Co-evolution of lexical meaning \& pragmatic use

### 2017-03-24 paper/paper-v3.tex
overhaul section 3 and 4; this version passed back to TB and RvR

The model of the previous section formalizes the evolutionary competition between different
pairs of lexical concepts and ways of using them. This section looks at a concrete case study
on scalar implicatures. Concretely, we will engage in the following formal thought experiment:
if a population of language users could freely combine different lexica with different
pragmatic strategies, what there the conditions under which the majority view of scalar
implicatures could have evolved?

Recall that the majority view is that scalar implicatures are non-lexicalized pragmatic
enrichments. Scalar implicature triggers like \emph{some}, \emph{warm} or \emph{may} are
semantically weak expressions for which logically stronger expressions are salient, e.g.,
\emph{all}, \emph{hot} or \emph{must}. For instance, {\em some} is entailed by {\em all}. If
the sentence ``Chris owns all of Johnny Cash's albums'' is true, then ``Chris owns some of
Johnny Cash's albums'' is also true. However, while weaker expressions such as {\em some} are
truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what
their use is normally taken to convey. Instead, the use of a less informative expression when a
more informative one could have been used can license a defeasible inference that stronger
alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). That is, a hearer who assumes
the speaker to be able and willing to provide all relevant information can infer that stronger
alternatives do not hold because the speaker used a weaker alternative instead. In this way,
``Chris owns some of Johnny Cash's albums'' is strengthened to convey that she owns {\em some
  but not all} albums. The upper bound, so the majority view, is a pragmatic inference but not
part of the conventional meaning. \mf{I wonder if this paragraph is really still necessary
  after the introduction?}

@@

To fill the model from Section~\ref{sec:model} with life, we need to specify the sets of
states, messages and lexica. Additionally, we want to explore the effects of a learning bias in
favor of simple lexical concepts. One way of motivating and formalizing such a bias is
introduced below.

\subsubsection{States, messages, concepts and Lexica}

Consider a state space with three states $\States = \set{\snone, \ssome, \sall}$ and think of
it as a partition of possible worlds into cells where none, some or all of the $A$s are $B$s,
for some arbitrary fixed predicates $A$ and $B$. Eight concepts can be distinguished based on
their truth or falsity in three world states, six of which are not contradictory or
tautological.

A lexicon $L$ is a mapping $\Messgs \rightarrow C$ from messages to concepts. With three
messages there are $6^3 = 216$ possible lexica. Some assign the same concept to more than one
message and others lexicalize the same concepts but associate them with different messages. Out
of these possible lexica, three kinds are of particular relevance. First, lexica that assign
the same concept to more than one message. Such lexica lack in expressivity but may be favored
by particular learning biases nonetheless. Second, lexica that conventionalize upper-bounds to
realize a (quasi-)partition of the state space. Finally, lexically that do not lexicalize an
upper bound but allow it to be conveyed pragmatically due to the presence of a stronger lexical
item. There are six lexica of the second kind and six of the third. The following three lexica
exemplify each kind:
\begin{align*}
  \begin{blockarray}{lccc}
    & & \underline{\Lall} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
     \mygray{\snone}  & 0 & 0 & 0\\
     \mygray{\ssome}  & 0 & 0 & 0\\
    \mygray{\sall}   & 1 & 1 & 1\\
    \end{block}
% \\ & & P(\Lall)\rlap{ $\propto 9$} & \\
  \end{blockarray} & &
 \begin{blockarray}{lccc}
    & & \underline{\Lbound} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 0 & 1\\
    \end{block}
%  \\& & P(\Lbound)\rlap{ $\propto 96$} & \\
  \end{blockarray} & &
  \begin{blockarray}{lccc}
    & & \underline{\Llack} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 1 & 1\\
    \end{block}
%   \\& & P(\Llack)\rlap{ $\propto 48$} & \\
  \end{blockarray}
\end{align*}
Lexicon $\Lall$ is clearly bad for communication: all message and interpretation choices will
be equally likely for all types; no information will be conveyed with it. Users of $\Lbound$
need not resort to pragmatic reasoning to convey an upper-bound with weak scalar
expressions. Instead, this information is already provided by the semantics of their lexicon,
giving them an advantage in terms of expressivity. Types with such lexica may be expected to
generally achieve higher communicative success than pragmatic users with a lexicon of type
$\Llack$ (see Section~\ref{sec:languages+use}).

Recall that types are a combination of a lexicon and a manner of language use. We analyze the
model's predictions in populations of types with one of the two behaviors introduced earlier;
literal or pragmatic. The former correspond to level $0$ reasoners and the latter to ones of
level $1$. Accordingly, we consider a total of $432$ types. Six are variants of pragmatic
language users with $\Llack$-like lexica. These will be our target types, because they
represent lexica and language use that conform to the majority view of scalar
implicatures. Twelve types are either literal or pragmatic speakers with lexical of the
$\Lbound$ kind. These will be the target types main competitors in evolutionary competition.

@@

There is a growing effort to develop empirically testable representational languages that allow
for the measure of semantic complexity. For instance, so-called {\em languages of thought}
(LOT) have been put to test in various rational probabilistic models that show encouraging
results (see, e.g., \citealt{katz+etal:2008, piantadosi+etal:underreview,
  piantadosi+etal:2012a} and \citealt{piantadosi+jacobs:2016} for recent discussion). At its
core, a LOT defines a set of operations and composition rules from which concepts can be
derived. As a first approximation and for the sake of concreteness, we follow this approach to
make the preference of learners for simpler semantic representations precise and define the
complexity of a representation as a function of its derivation cost in a weighed generative
LOT.

\begin{table}
  \centering
  \begin{align*}
    C & \rightarrow_2 C \wedge C 
    & 
    X & \rightarrow_1 \set{A,B} \\
    C & \rightarrow_2 \neg C 
    & 
    X & \rightarrow_1 X \cap X \\
    C & \rightarrow_1 X \subseteq X
    & 
    X & \rightarrow_1 X \cup X \\
    C & \rightarrow_1 X \neq \emptyset \\
    C & \rightarrow_1 X = \emptyset     
  \end{align*}
  \caption{Toy grammar in a set-theoretic LOT with weighted rules.}
  \label{tab:grammar}
\end{table}

Our toy grammar of concepts is given in Table~\ref{tab:grammar}. This grammar uses basic
set-theoretic operations to form expressions which can be evaluated as true or false in states
$\snone$, $\ssome$ or $\sall$ from above. Applications of generative rules have a cost attached
to them. Here we simply assume that Boolean combinations of concepts are more complex than
atomic concepts and that otherwise each rule application adds the same cost
unit. Table~\ref{tab:concepts} lists all six concepts relevant here, their truth conditions and
the simplest formula that expresses this concept from the grammar in Table~\ref{tab:grammar}.

\begin{table}
  \centering
\begin{center}
  \begin{tabular}{lccclc}
    \toprule
    intuitive name
    & \snone
    & \ssome
    & \sall
    & least complex formula
    & complexity
    \\ \midrule
    ``all''
    & 0
    & 0
    & 1
    & $A \subseteq B$
    & $3$
    \\
    ``some but not all''
    & 0
    & 1
    & 0
    & $A \cap B \neq \emptyset \wedge A \neq \emptyset$
    & $8$
    \\    
    ``some''
    & 0
    & 1
    & 1
    & $A \cap B \neq \emptyset$
    & $4$
    \\
    ``none''
    & 1
    & 0
    & 0
    & $A \cap B = \emptyset$
    & $4$
    \\
    ``none or all''
    & 1
    & 0
    & 1
    & $\neg(A \cap B \neq \emptyset \wedge A \neq \emptyset)$
    & $10$
    \\
    ``not all''
    & 1
    & 1
    & 0
    & $\neg (A \subseteq B)$
    & $5$
    \\
    \bottomrule
  \end{tabular}
\end{center}
\caption{Available concepts and their minimal derivation length}
\label{tab:concepts}
\end{table}

The complexity measure for lexical concepts from Table~\ref{tab:concepts} is used to define a
learning bias that favor simpler concepts over more complex ones. The prior probability of a
type is just the prior probability of its lexicon. The prior of a lexicon is a function of the
complexity of the lexical representations in its image set. Lexica with simpler concepts have a
higher prior. One simple way of defining such priors over lexica (and thereby types) is:
\begin{align*}
  P(L)  & \propto \prod_{c \in Im(L)} P(c)   \ \text{, with} & 
  P(c) & \propto \max_{c'}Compl(c') - Compl(c) + 1\,,
\end{align*}
where $Compl(c)$ is the complexity of the minimal derivation length of $c$ according to the
LOT-grammar (see Table~\ref{tab:concepts}). There are many ways to define priors over lexica
(see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a}) but the key assumption here,
common to all of them, is that simple representational expressions should be favored over more
complex ones. We should stress that these details -- from the generative grammar to its
complexity measure -- are to be regarded as one convenient operationalization of one general
approach to explicating learning biases; this is not a commitment that this general approach is
necessarily superior or that, within it, this particular instrumentalization is the single most
plausible.

@@

Recall that there are three parameters: soft-max parameter $\lambda$ affects how strongly
pragmatic speakers favor more informative messages; the bottleneck size $k$ influences how
faithfully learners can identify their teacher type; $l$ defines the learners disposition
towards choosing the most likely teacher type from the posterior distribution. We expect that
competitor types (types with lexica of the kind $\Lbound$) have a fitness advantage over target
types (pragmatic agents with lexica of the kind $\Llack$), especially for very low levels of
$\lambda$. Selection based on fitness alone may therefore not lead to prevalence of target
types in the population. On the other hand, lexica of type $\Llack$ are simpler than those of
type $\Lbound$ by the postulated measure from above. This may make them more likely to be
adopted by learners, especially when $k$ is low so that teacher types are relatively
indistinguishable based on their behavior in general and when $l$ is high. Still, lexica of the
kind $\Lall$ are in turn more likely \emph{a priori} than lexica of the kind
$\Llack$. Simulation results will shed light on the question whether target types can emerge
and for which parameter constellations.

To better understand the workings a pressure towards communicative efficiency and a pressure
towards learnability, we look at the behavior of the replicator and mutator step first in
isolation, and then in combination. All simulation runs are initialized with an arbitrary
distribution over types, constituting a population's first generation. All reported results are
the outcome of 50 update steps for each different constellation of parameter values. This
ensures that these outcomes correspond to developmental plateaus in which change is, if not
absent, then at least very slow, In other words, even if the resulting states do not correspond to
an eventual attracting state, they characterize long almost stationary states in which the
system remains for a very long time. As specified in \S\ref{sec:learnability}, the mutation
matrix $Q$ can be obtained by considering all possible state-message sequences of length
$k$. Given that this is intractable for large $k$, the sets of data learners are exposed to are
approximated by sampling $250$ $k$-length sequences from each type's production probabilities.

\subsubsection{Replication only: selection based on expressivity and communicative success} 
\begin{figure}
\centering
\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a pressure for expressivity. The right-most plots zoom in on only the proportion of pragmatic $\Llack$ types.}
\label{fig:only-R}
\end{figure}

As expected, selection based on communicative success is sensitive to $\lambda$ as it
influences signaling behavior. This is showcased in Figure \ref{fig:only-R}, which depicts the
proportion of target types in $5$ independent populations after $50$ generations, as well as
that of the type with the highest proportion in a given population, the incumbent. With low
$\lambda$ many types have very similar behavior, so that evolutionary selection lacks grip and
becomes very slow. The result is a very long transition with near stagnancy in a rather
homogeneous population with many types. Conversely, higher $\lambda$ promotes more rational
linguistic behavior, widening the gap in expressivity between types and promoting more
homogeneous populations. As suggested by Figure \ref{fig:only-R}, the incumbent in most
populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure only for
expressivity does not lend a justification to the systematic prevalence of a lack of
upper-bounds under any $\lambda$-value. For instance, in $1000$ independent populations with
$\lambda = 20$ pragmatic $\Llack$ types only made up $11$ cases of incumbency, corresponding to
a mean proportion of $0.003$ across populations. By contrast, $913$ incumbents were
$\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types
($459$), corresponding to a mean proportion of about $.48$ taken
together. %Across types and populations incumbents make up $.50$ of the mean outcome.

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a pressure for learnability ($\lambda = 20, k = 5$). The types' prior probability is shown in the right-most plot.}
\label{fig:only-M}
\end{figure}

The effect of iterated learning without a pressure for expressivity using either posterior
sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$) is shown in
Figure \ref{fig:only-M} together with the prior over types. The prior shows that while users of
$\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are
nevertheless more advantaged than others, such as $\Lbound$, in virtue of the relatively simple
semantics they conventionalize. Crucially, $\Llack$ enables its users to convey each state with
a single message when combined with pragmatic reasoning provided sufficiently high
$\lambda$. This makes it less likely to be confused with other types if the learning data is
not too sparse ($k \geq 5$). Put differently, learners have a higher propensity to infer
pragmatic $\Llack$ over types that might be hard or impossible to distinguish it based on overt
linguistic behavior, such as $\Lbound$. Moreover, $\Llack$ is also less likely to be confused
with types with different observable behavior because its pragmatic use approximates a
one-to-one form-meaning mapping. As a consequence, a stronger propensity to maximize the
posterior increases their proportion in the population.

However, in contrast to a pressure only for expressivity with high $\lambda$, learnability
alone does not succeed in selecting for a single prevalent type. Indeed, all six target types
tend to coexist at roughly equal proportion. Each is passed on to next generation with the same
faithfulness and, differently from a pressure for communicative success, they do not stand in
competition with each other. As implicit in Figure \ref{fig:only-M}, in $1000$ independent
populations all incumbents were pragmatic $\Llack$ users provided sufficiently high $\lambda$,
with each reaching approximately the same proportion of users in the population. As with
expressivity only, low values of $\lambda$ make the differences in observable behavior across
types less pronounced and therefore reflect the learners' inductive bias more faithfully,
favoring functionally deficient but a priori preferred types such as those that use $\Lall$. A
pressure for learnability alone may consequently lead to a spread of communicatively suboptimal
types that are easier to learn. In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$
independent populations had users of $\Lall$ as incumbents.

@@

Pressures for communicative success or learnability are not sufficient on their own to have a
single target type dominate the population. When pressured toward expressivity, the slight
communicative advantage of $\Lbound$ users leads to the its prevalence. When pressured toward
learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
complex alternatives such as $\Lbound$. However, on its own learnability does not foment the
propagation of these types across the population. What is more, it leads to heterogeneous
populations that foster an equal share of types of a kind, such as different pragmatic $\Llack$
types.

\begin{figure}
\centering
\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig3-r+m}
\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under both pressures ($k = 5$).}
\label{fig:rmd}
\end{figure}

Figure \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of $\lambda$ and $l$ values. These results show that an inductive learning bias for simpler semantics in tandem with functional pressure can lead to the selection of $\Llack$-like semantics paired with pragmatic language use. The proportion of these types increases with $\lambda$ and $l$. Pressure towards expressivity magnifies the effects of iterated learning and dampens the proliferation of types of a kind that are equal in expressivity {\em and} learnability. A pressure towards learnability favors the transmission of simpler semantics and thereby makes types that exploit pragmatic reasoning to strengthen lexical meaning more likely to be adopted over alternatives that do not. %In short, neither learnability nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions in predominantly monomorphic populations.

As before, low $\lambda$ and $l$ lead to the incumbency of communicatively suboptimal types that are a priori favored, such as $\Lall$. An increase in $\lambda$ leads to the selection and incumbency of pragmatic $\Llack$ but does not lead to monomorphic populations if learners sample from the posterior. Finally, a combination of sufficiently rational signaling behavior with some tendency for posterior maximization leads to increasingly monomorphic populations of types that convey upper-bounds pragmatically. The difference between the mean of the highest pragmatic $\Llack$-like type in $1000$ independent populations and the highest proportion of other types across $\lambda$ and $l$ values is shown in Figure \ref{fig:diff}.

>>>>

The model of the previous section formalizes the evolutionary competition between different
pairs of lexical concepts and ways of using them. This section looks at a case study on scalar
implicatures. It engages in a following formal thought experiment to address the question: if a
population of language users could freely combine different lexica with different pragmatic
strategies, what are conditions under which the majority view of scalar implicatures could have
evolved?

Recall that the majority view is that scalar implicatures are non-lexicalized pragmatic
enrichments. Scalar implicature triggers like \emph{some}, \emph{warm} or \emph{may} are
semantically weak expressions for which logically stronger expressions are salient, e.g.,
\emph{all}, \emph{hot} or \emph{must}. For instance, {\em some} is entailed by {\em all}. If
the sentence ``Chris owns all of Johnny Cash's albums'' is true, then ``Chris owns some of
Johnny Cash's albums'' is also true. However, while weaker expressions such as {\em some} are
truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what
their use is normally taken to convey. Instead, the use of a less informative expression when a
more informative one could have been used can license a defeasible inference that stronger
alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). In this way, ``Chris owns some
of Johnny Cash's albums'' is strengthened to convey that she owns {\em some but not all}
albums. According to the majority view, this is a pragmatic inference, not part of the
conventional meaning.

@@

To fill the model from Section~\ref{sec:model} with life, we need to specify the sets of
states, messages and lexica (Section~\ref{sec:stat-mess-conc}). Additionally, we want to
explore the effects of a learning bias in favor of simple lexical concepts. One way of
motivating and formalizing such a bias is introduced below in
Section~\ref{sec:an-induct-learn}.

\subsubsection{States, messages, concepts and Lexica} 
\label{sec:stat-mess-conc}

Consider a state space with three states $\States = \set{\snone, \ssome, \sall}$ and think of
it as a partition of possible worlds into cells where none, some or all of the $A$s are $B$s,
for some arbitrary fixed predicates $A$ and $B$. Eight concepts can be distinguished based on
their truth or falsity in three world states, six of which are not contradictory or
tautological (see Table~\ref{tab:concepts} below).

A lexicon $L$ is a mapping $\Messgs \rightarrow C$ from messages to concepts. With three
messages there are $6^3 = 216$ possible lexica. Some assign the same concept to more than one
message and others lexicalize the same concepts but associate them with different messages. Out
of these possible lexica, three kinds are of particular relevance. First, lexica that assign
the same concept to more than one message. Such lexica lack in expressivity but may be favored
by particular learning biases nonetheless (see below). Second, lexica that conventionalize
upper-bounds to realize a one-to-one mapping of messages to states. Finally, lexically that do
not lexicalize an upper bound but allow it to be conveyed pragmatically due to the presence of
a stronger lexical item. There are six lexica of the second kind and six of the third. The
following three lexica exemplify each kind:
\begin{align*}
  \begin{blockarray}{lccc}
    & & \underline{\Lall} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
     \mygray{\snone}  & 0 & 0 & 0\\
     \mygray{\ssome}  & 0 & 0 & 0\\
    \mygray{\sall}   & 1 & 1 & 1\\
    \end{block}
% \\ & & P(\Lall)\rlap{ $\propto 9$} & \\
  \end{blockarray} & &
 \begin{blockarray}{lccc}
    & & \underline{\Lbound} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 0 & 1\\
    \end{block}
%  \\& & P(\Lbound)\rlap{ $\propto 96$} & \\
  \end{blockarray} & &
  \begin{blockarray}{lccc}
    & & \underline{\Llack} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 1 & 1\\
    \end{block}
%   \\& & P(\Llack)\rlap{ $\propto 48$} & \\
  \end{blockarray}
\end{align*}
Lexicon $\Lall$ is clearly bad for communication: all message and interpretation choices will
be equally likely for all types; no information about the observed world state will be conveyed
by its users. In contrast, users of $\Lbound$ can communicate world states perfectly, no matter
whether they are literal or pragmatic users. Users of $\Llack$ can also communicate information
about the actual world state but need pragmatic language use to approximate a one-to-one
mapping between message use and states (see Section~\ref{sec:languages+use}).

Recall that types are a combination of a lexicon and a manner of language use. We analyze the
model's predictions in populations of types with one of the two behaviors introduced earlier;
literal or pragmatic. The former correspond to level $0$ reasoners and the latter to level
$1$. Accordingly, we consider a total of $432$ types. Six are variants of pragmatic language
users with $\Llack$-like lexica. We refer to these as \emph{target types}, because they
represent lexica and language use that conform to the majority view of scalar
implicatures. Twelve types are either literal or pragmatic speakers with lexical of the
$\Lbound$ kind. We refer to these as \emph{competitor types}, because they are expected to be
the target types' main competitors in evolutionary competition.

@@

There is a growing effort to develop empirically testable representational languages that allow
for the measure of semantic complexity. For instance, so-called {\em languages of thought}
(LOTs) have been put to test in various rational probabilistic models that show encouraging
results (see, e.g., \citealt{katz+etal:2008, piantadosi+etal:underreview,
  piantadosi+etal:2012a} and \citealt{piantadosi+jacobs:2016} for recent discussion). At its
core, a LOT defines a set of operations and composition rules from which concepts can be
derived. As a first approximation and for the sake of concreteness, we follow this approach to
motivate and formalize a preference of learners for simpler semantic representations
\citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a,
  kirby+etal:2015,piantadosi+etal:underreview}. A concept's complexity is a function of its
derivation cost in a weighed generative LOT.

\begin{table}
  \centering
  \begin{align*}
    C & \rightarrow_2 C \wedge C 
    & C & \rightarrow_2 \neg C  \\ 
    C & \rightarrow_1 X \subseteq X &
    C & \rightarrow_1 X \neq \emptyset &
    C & \rightarrow_1 X = \emptyset  \\
    X & \rightarrow_1 \set{A,B} & 
    X & \rightarrow_1 X \cap X & 
    X & \rightarrow_1 X \cup X    
  \end{align*}
  \caption{Toy grammar in a set-theoretic LOT with weighted rules.}
  \label{tab:grammar}
\end{table}

Our toy grammar of concepts is given in Table~\ref{tab:grammar}. This grammar uses basic
set-theoretic operations to form expressions which can be evaluated as true or false in states
$\snone$, $\ssome$ or $\sall$ from above. Applications of generative rules have a cost attached
to them. Here we simply assume that the formation of Boolean combinations of concepts incurs 2
cost units, while all other rule applications incur only 1 cost unit. Table~\ref{tab:concepts}
lists all six concepts relevant here, their truth conditions and the simplest formula that
expresses this concept from the grammar in Table~\ref{tab:grammar}.

\begin{table}
  \centering
\begin{center}
  \begin{tabular}{lccclc}
    \toprule
    intuitive name
    & \snone
    & \ssome
    & \sall
    & least complex formula
    & complexity
    \\ \midrule
    ``all''
    & 0
    & 0
    & 1
    & $A \subseteq B$
    & $3$
    \\
    ``some but not all''
    & 0
    & 1
    & 0
    & $A \cap B \neq \emptyset \wedge A \neq \emptyset$
    & $8$
    \\    
    ``some''
    & 0
    & 1
    & 1
    & $A \cap B \neq \emptyset$
    & $4$
    \\
    ``none''
    & 1
    & 0
    & 0
    & $A \cap B = \emptyset$
    & $4$
    \\
    ``none or all''
    & 1
    & 0
    & 1
    & $\neg(A \cap B \neq \emptyset \wedge A \neq \emptyset)$
    & $10$
    \\
    ``not all''
    & 1
    & 1
    & 0
    & $\neg (A \subseteq B)$
    & $5$
    \\
    \bottomrule
  \end{tabular}
\end{center}
\caption{Available concepts and their minimal derivation cost}
\label{tab:concepts}
\end{table}

The complexity measure for lexical concepts from Table~\ref{tab:concepts} is used to define a
learning bias that favors simpler concepts over more complex ones. The prior probability of a
type is just the prior probability of its lexicon. The prior of a lexicon is a function of the
complexity of the lexical representations in its image set. Lexica with simpler concepts have a
higher prior. One simple way of defining such priors over lexica (and thereby types) is:
\begin{align*}
  P(L)  & \propto \prod_{c \in Im(L)} P(c)   \ \text{, with} & 
  P(c) & \propto \max_{c'}Compl(c') - Compl(c) + 1\,,
\end{align*}
where $Compl(c)$ is the complexity of the minimal derivation cost of $c$ according to the
LOT-grammar (see Table~\ref{tab:concepts}). This construal assigns the highest probability the
above example for a lexicon of type $\Lall$, which only uses the simplest concept ``all'' for
all messages. Lexica of type $\Llack$ are less likely, but more likely than $\Lbound$.

There are many ways to define priors over lexica (see, e.g., \citealt{goodman+etal:2008,
  piantadosi+etal:2012a}) but the key assumption here, common to all of them, is that simple
representational expressions should be favored over more complex ones. We should stress that
these details -- from the generative grammar to its complexity measure -- are to be regarded as
one convenient operationalization of one general approach to explicating learning biases; this
is not a commitment that this general approach is necessarily superior or that, within it, this
particular instrumentalization is the single most plausible.

@@

Recall that there are three parameters: soft-max parameter $\lambda$ affects how strongly
pragmatic speakers favor messages that appear best from their subjective point of view; the
bottleneck size $k$ influences how faithfully learners can identify their teacher type; $l$
defines the learners' disposition towards choosing the most likely teacher type from the
posterior distribution. We expect that competitor types (types with lexica of the kind
$\Lbound$) have a fitness advantage over target types (pragmatic agents with lexica of the kind
$\Llack$), especially for very low levels of $\lambda$. Selection based on fitness alone may
therefore not lead to prevalence of target types in the population. On the other hand, lexica
of type $\Llack$ are simpler than those of type $\Lbound$ by the postulated measure from
above. This may make them more likely to be adopted by learners, especially when $k$ is low so
that different teacher types are relatively indistinguishable based on their behavior and when
$l$ is high. Still, lexica of the kind $\Lall$ are in turn more likely \emph{a priori} than
lexica of the kind $\Llack$. Simulation results will shed light on the question whether target
types can emerge and for which parameter constellations.

To better understand the workings of a pressure towards communicative efficiency and a pressure
towards learnability, we look at the behavior of the replicator and mutator step first in
isolation, and then in combination. All simulation runs are initialized with an arbitrary
distribution over types, constituting a population's first generation. All reported results are
the outcome of 50 update steps. These outcomes correspond to developmental plateaus in which
change is, if not absent, then at least very slow. In other words, even if the resulting states
do not correspond to an eventual attracting state, they characterize long almost stationary
states in which the system remains for a very long time. As specified in
\S\ref{sec:learnability}, the mutation matrix $Q$ can be obtained by considering all possible
state-message sequences of length $k$. Given that this is intractable for large $k$, the sets
of data learners are exposed to are approximated by sampling $250$ $k$-length sequences from
each type's production probabilities.

\subsubsection{Replication only: selection based on expressivity and communicative success}

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other
  than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a
  pressure for expressivity. The right-most plots zoom in on only the proportion of pragmatic
  $\Llack$ types. \mf{not sure if ``incumbent'' is the right term here; I believe it means
    ``someone who is already there and established'' but we mean ``type with the highest
    proportion''; I used ``majority type'' in the main text; maybe change the plot?}}
\label{fig:only-R}
\end{figure}

As expected, selection based on communicative success is sensitive to $\lambda$ as it
influences signaling behavior, which in turn determines communicative success. This is
showcased in Figure \ref{fig:only-R}, which depicts the proportion of target types in $5$
independent populations after $50$ replicator steps. The plot also shows the proportion of the
\emph{majority type}, i.e., the type with the highest proportion in the final population. With
low $\lambda$ many types have very similar behavior, so that evolutionary selection lacks grip
and becomes very slow. The result is a very long transition with near stagnancy in a rather
homogeneous population with many types. Conversely, higher $\lambda$ promotes more rational
linguistic behavior, widening the gap in expressivity between types and promoting more
homogeneous populations. As suggested by Figure \ref{fig:only-R}, the incumbent in most
populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure only for
expressivity does not lead to a prevalence of target types under any $\lambda$-value. For
instance, in $1000$ independent populations with $\lambda = 20$ only had $11$ cases in which
the target type was the majority type, corresponding to a mean proportion of $0.003$ across
populations. By contrast, in $913$ cases the majority types had $\Lbound$ with close to an even
share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion
of about $.48$ taken together.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other
  than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a
  pressure for learnability ($\lambda = 20, k = 5$). The types' prior probability is shown in
  the right-most plot. \mf{explain the topmost part of the right plot?!}}
\label{fig:only-M}
\end{figure}

The effect of iterated learning without a pressure for expressivity using either posterior
sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$) is shown in
Figure \ref{fig:only-M} together with the prior over types. The prior shows that while users of
$\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are
nevertheless more advantaged than others, such as $\Lbound$, in virtue of the relatively simple
semantics they conventionalize (see Section~\ref{sec:an-induct-learn}). Crucially, $\Llack$
enables its users to convey each state with a single message when combined with pragmatic
reasoning provided sufficiently high $\lambda$. This makes it less likely to be confused with
other types if the learning data is not too sparse ($k \geq 5$). Put differently, learners have
a higher propensity to infer pragmatic $\Llack$ when the teacher's type produces very similar
data, such as when using $\Lbound$. Moreover, $\Llack$ is also less likely to be confused with
types with different observable behavior because its pragmatic use approximates a one-to-one
form-meaning mapping. As a consequence, a stronger propensity to maximize the posterior
increases their proportion in the population.

However, in contrast to a pressure only for expressivity with high $\lambda$ (see
Figure~\ref{fig:only-R}), learnability alone does not succeed in selecting for a single
prevalent type. Indeed, all six target types tend to coexist at roughly equal proportion. Each
is passed on to the next generation with the same faithfulness and, differently from a pressure
for communicative success, they do not stand in competition with each other. In $1000$
independent populations all majority types were target types provided, sufficiently high
$\lambda$ \mf{don't understand this proviso: what $\lambda$ values were used in the 1000
  runs? what's sufficiently high?}, with each reaching approximately the same proportion of users in the population. As
with expressivity only, low values of $\lambda$ make the differences in observable behavior
across types less pronounced and therefore reflect the learners' inductive bias more
faithfully, favoring functionally deficient but a priori preferred types such as those that use
$\Lall$. A pressure for learnability alone may consequently lead to a spread of communicatively
suboptimal types that are easier to learn. In the extreme, when $l = 1$ and $\lambda = 1$ all
of $1000$ independent populations had users of $\Lall$ as incumbents.

@@

Pressures for communicative success or learnability are not sufficient on their own to have a
single target type dominate the population. When pressured for expressivity, the slight
communicative advantage of $\Lbound$ users leads to the its prevalence. When pressured for
learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
complex alternatives such as $\Lbound$. However, on its own learnability does not foment the
propagation of a single target type across the population.

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig3-r+m}
\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under both pressures ($k = 5$).}
\label{fig:rmd}
\end{figure}

Figure \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of
$\lambda$ and $l$ values. These results show that an inductive learning bias for simpler
semantics in tandem with functional pressure can lead to the selection of a single target
type. The proportion of a single majority target type increases with $\lambda$ and
$l$. Pressure for communicative success magnifies the effects of iterated learning and dampens
the proliferation of types of a kind that are equal in expressivity {\em and} learnability. A
pressure towards learnability favors the transmission of simpler semantics and thereby selects
for pragmatic language use.

As before, low $\lambda$ and $l$ lead to the prevalence of communicatively suboptimal types
that are a priori favored, such as $\Lall$. An increase in $\lambda$ leads to the selection of
target types but does not lead to monomorphic populations if learners sample from the
posterior. Finally, a combination of high $\lambda$ and $l$, leads to increasing proportions of
a single majority target type. The difference between the mean of the highest pragmatic
$\Llack$-like type in $1000$ independent populations and the highest proportion of other types
across $\lambda$ and $l$ values is shown in Figure \ref{fig:diff}.

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth,height=10cm,keepaspectratio]{./plots/fig4-incumbents-difference}
\caption{Difference between mean proportion of highest pragmatic $\Llack$ type and highest
  other type in $1000$ independent populations after $50$ generations under both pressures ($k
  = 5$). \tb{I will run more simulations with higher $\lambda$ and $l$. I'm pretty sure the
    difference still increases.} \mf{I think it's okay like that} \mf{the coloring is not that
    easy to decipher here; would a two-dimensional surface be possible and look better?}}
\label{fig:diff}
\end{figure}
